Query Optimization for Dynamic Imputation

### 2017-05-15 paper/rebuttal.tex

\section{Overall changes}
\begin{itemize}[leftmargin=*]
\item \resp{
    Several reviewers commented on redundant figures, excessive whitespace, and
    formatting, so we've compressed many of the figures and removed whitespace to make room for additional material.
}

\item \resp{
    We have improved our system since the paper was submitted by adding imputation for categorical variables.
    This has a minor impact on the experimental results relative to the original submission.
    It improves the ability of the decision tree model to perform imputation by making these variables available to the model.
  }

\item \resp{We extended the original paper with two additional popular imputation techniques: mean value imputation and hot-deck imputation.

    \begin{sloppypar}
We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation}).
\end{sloppypar}
}

\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}). We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers}) and
explored planning times for more complex queries.}

@@

\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database (\Cref{sec:implementation}).}
\end{itemize}

  \section{Reviewer 1}
  \begin{itemize}[leftmargin=*]
  \item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.

    \resp{
      \begin{sloppypar}
    We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation.
    These methods are in addition to the original imputation method which is based on decision trees.
    The decision tree model incorporates correlations between attributes, so should be able to take advantage of any functional dependencies to make better predictions.
    Interpolation-based imputation is useful for data where the ordering of tuples is significant, like time series, but is not well suited for survey data where the ordering of tuples is arbitrary.
  \end{sloppypar}

@@

   Extending a rule-based optimizer to perform dynamic imputation would certainly be an interesting idea. However, the planning algorithm presented benefits from the tight
   interaction between traditional query planning and imputation operator placement. The latter can influence aspects such as join ordering and cardinality estimates for subplans.
   Incorporating these into a rule-based system would be possible, but we believe the plan selected through this approach will be suboptimal, compared to that chosen through
   our integrated planning approach.
   }

  \item D3. Is there any way for the user to know what error to expect based on the $\alpha$ value? Moreover, is its value somehow proportional to the error (e.g., $\alpha=0.5$ would give roughly double the error of $\alpha=0.25$)? Or should the user simply try multiple execution with different values of $\alpha$ until the desired results are given?

    \resp{In our intended usage, the user starts by running with $\alpha = 0$ and increases $\alpha$ if the query takes too long to complete.

     We agree that having a way of providing an estimate of error, based on $\alpha$, would be both useful and interesting to develop.
     Providing error estimates further requires a reasonable notion of ground-truth. For evaluation, we've used the values imputed on the 
     tables when applying whole-table imputation. Based on our evaluation, higher values of $\alpha$ correspond to increased error, however, extending $\alpha$ to be predictive is not something we implemented in the current system.
     }

@@

  \item D8. Are there impute implementation that are non-blocking? Or do they always have to perform multiple passes over the data?

    \resp{Yes. For example, in time series data, it is reasonable to fill the most recent non-null value forward. This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments. We have extended the experiments section to consider a simple non-blocking imputation method relevant to our dataset: mean value imputation. This is a common technique, which fills in  missing values
    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
    the data before imputing (\Cref{subsubsec:mean}).}

  \item D9. Additional experiments I suggest to include:
    -- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
    -- Try different values for $\alpha$ (not just 0 and 1).
    -- Show runtime of enumeration algorithm as queries get more complex.

    \resp{
      It is straightforward to compute the number of queries before we lose our advantage.
      For example, performing full imputation on the ACS dataset takes 355 minutes (Sec. 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation. We've made this point explicit in the ACS experiment section.

### 2017-05-15 paper/introduction.tex

A simpler approach might drop all rows with any missing values, which can not only introduce
bias into the results, but also result in discarding much of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
carefully plans the placement of imputation or row-drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the
Pareto-optimal frontier of plans that trade imputation cost for result quality,
and allow the analyst to specify their desired plan from this frontier.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazero{} times faster than
imputing all missing values and then running queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaone{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

@@

These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

A simpler approach might drop all rows with any missing values, which can not only introduce
bias into the results, but also result in discarding much of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
carefully plans the placement of imputation or row-drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the
Pareto-optimal trade-offs between imputation cost for result quality,
and allow the analyst to specify their desired plan from this frontier.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazero{} times faster than
imputing all missing values and then running queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaone{}
percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

@@

These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto-optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-05-15 paper/main.tex
Add sentence on tree runtimes

@@

@@

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazero{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
\end{abstract}

>>>>

\setlist[itemize]{leftmargin=2em,itemsep=-0.2em}

@@

\newcommand{\runtimetreelowzero}{75 ms}
\newcommand{\runtimetreehighzero}{1 second}
\newcommand{\runtimetreelowone}{12 ms}
\newcommand{\runtimetreehighone}{19 ms}
\newcommand{\runtimetreelowbase}{6.5 seconds}
\newcommand{\runtimetreehighbase}{27 seconds}

@@

\newcommand{\runtimemeanlow}{12 ms}
\newcommand{\runtimemeanhigh}{25 ms}
\newcommand{\runtimemeanlowbase}{24 ms}
\newcommand{\runtimemeanhighbase}{44 ms}
\newcommand{\runtimehotdecklow}{12 ms}
\newcommand{\runtimehotdeckhigh}{29 ms}
\newcommand{\runtimehotdecklowbase}{21 ms}
\newcommand{\runtimehotdeckhighbase}{92 ms}

\newcommand{\smapehotdecklow}{0}
\newcommand{\smapehotdeckhigh}{3}
\newcommand{\smapehotdeckhighoutlier}{24}

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazero{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
\end{abstract}
